Implementing \& Evaluating Space Efficient Algorithms for Detecting Large Neighbourhoods in Graph Streams

### 2020-05-03 paper/src/paper.tex
Write up almost done

\usepackage[ruled, vlined, linesnumbered]{algorithm2e}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{changepage}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{fancyhdr} % Header

@@

\renewcommand{\headrulewidth}{0pt}
\newcommand{\ie}{\textit{i.e.} }
\newcommand{\nats}{\mathbb{N} }
\newcommand{\horizontalline}{\newline\vspace{.3cm}\hfill\makebox[.5\linewidth]{\rule{.5\textwidth}{0.4pt}}\hfill\vspace{.1cm}}
% title
\title{Implementing \& Evaluating Space Efficient Algorithms for Detecting Large Neighbourhoods in Graph Streams}
\author{Dom Hutchinson}
\date{\today}
\maketitle

% Header
\pagestyle{fancy}
\fancyhead[L]{Dom Hutchinson (1701111)}
\fancyhead[C]{}
\fancyhead[R]{\today}

\chapter{Background}

@@

\par When designing algorithms time efficiency is generally viewed with primary importance, over space efficiency. However, if the memory used for an algorithm exceeds the RAM space allocated to it during execution then virtual memory would have to be used. Virtual memory has a much longer read/write time than RAM meaning that an algorithm's run-time will be much greater in this scenario. Thus, algorithms which are designed to work with large data sets space efficiency should be viewed as having primary importance.
\par In the early 2000s companies began to realise that data could be used as a commodity, and with this, the amount of data being collected has surged in the years since. In this same time period, we have not seen the same surge in the amount of RAM typical computers have. Thus people wishing to capitalise on this increase in available data need to give greater thought to the space efficiency of the algorithms they are using. Otherwise, run-times could become unusable.
\par Social networks are good sources of data about people. This data can be leveraged in many ways, notably within advertising. Advertisements which are targeted at people with certain observed behaviours have been shown to have more than twice the click-through rate of standard online advertisements \cite{targetedAds}. This has motivated the development of techniques for assessing the behaviours of a person just from their online presence.
\par One technique is to look at who a person is connected to and then to use observations about these connections to infer the behaviours of said person. In turn, we have seen the rise of the ``Social Media Influencer". These are people who are deemed to have a large online following and, moreover, a following which they are able to influence. Advertisers will approach influencers with deals by which the influencer will promote a product to their audience, hopefully driving sales.
\par The question for advertisers now becomes ``Who is an influencer?". To answer this we need to indentify people with large neigbourhoods in a network \& members of this network. We need to find the members of the neighbourhood (or at least a subset of them) in order to be able to assess what the demographics of the influencers audience are. This question can be generalised as the $\mathtt{Neighborhood\ Detection\ Problem}$. Which is the problem I shall discuss in this paper.
\par Although in my description of the Neighbourhood Detection Problem I motivate its relevance by referring to finding social media influencers, it is applicable to many other scenarios.
\begin{itemize}
	\item[-] Which items are most commonly bought \& which other items they are commonly bought with?
	\item[-] Which resources in a network are being accessed most \& by whom?
\end{itemize}
%TODO More
Answering these questions can be used to inform resource allocation in order to maximise a certain goal. Be that profit; or, network response speeds.
\section{Neighbourhood Detection Problem}
% Describe Graph streams, insertion-only & insertion-deletion
\par Graph Streams are a list of instructions for how to construct a graph and provide a dynamic way of representing graphs. In this paper, I shall discuss two types of graph streams: Insertion-Only Streams; and, Insertion-Deletion Streams. There are other types of graph streams that allow for additional operations such as updating vertex \& edge values, but these are not relevant to the problem being considering in this paper.
\par Insertion-Only graph streams are a list of pairs of vertices, representing endpoints of unweighted-undirected edges. Each instruction describes a new edge to be inserted into the graph \& potential up to two new nodes which are in the graph. There is no way to remove or update an edge in an Insertion-Only graph stream. We apply a limitation that no duplicate edges appear in an insertion-only stream and thus the order of instructions in an Inserionly-Only stream has no effect on the validity of the algorithms which shall be discussed.
\par Insertion-Deletion graph streams consist of a pair of vertices and a boolean. The vertices represent the endpoints of an unweighted-undirected edge \& the boolean describes whether this edge is being inserted or deleted from the graph. If it is an insertion instruction then we are adding the described edge to the graph; if it is a deletion instruction then we are removing it. Again we add a limitation that no edges are repeated \& that we will not receive a request to delete an edge from the graph which does not currently exist in the graph. This does, however, mean that the order of instructions in an Insertion-Deletion stream is important as the deletion of an edge must come after it's insertion.
\par The limitations placed here often occur naturally in systems. On social networks you are not able to follow someone you are already following, nor unfollow someone you are not following.
\par In spite of these limitations which are imposed it is still possible to construct almost any undirected-unweighted graph using the two graph streams described. Suppose you wish to analyse which devices in a network are making the most request. Here vertices represent devices \& edges represent requests. For this scenario, you would want to allow for multiple edges (say $n>2$) between the same pair of devices (say $\{x,y\}$). One solution would be to add $n$ new vertices $\{v_{x,y,1},\dots,v_{x,y,n}\}$ each with a label stating they are just intermediate nodes and not full devices. Creating edges from $x$ to each of $\{v_{x,y,1},\dots,v_{x,y,n}\}$ \& from $y$ to each of $\{v_{x,y,1},\dots,v_{x,y,n}\}$ allows for an interpretation of multiple occurrences of the edge $(x,y)$.
\par If a graph stream is ordered by time (which is common in real-life applications) then it represents a dynamic-evolution of the graph over time. Allowing the user to roll back the graph should they wish to make assessments during particular time periods, rather than just assess the final graph.
\par The algorithms which shall be discussed in this paper are designed to work with just a single pass of a graph stream. This is not only time-efficient as it does not require multiple reads of the stream, but means that should more information be gained after the algorithm is run it is easy to incorporate it into the algorithm if we store the final state of the data structures used. Single-pass algorithms mean we don't necessarily have to store the graph stream in a local file, we could simply send instructions from the system being modeled straight into the algorithm. This is useful when a graph stream potentially could take up multiple terra-bytes, but is unnecessary in most applications and can mean it is harder to verify results later.
\horizontalline
% Build to problem
\par The $\mathtt{High\ Degree\ Detection\ Probem}$ is a problem in graph theory where we seek to find a vertex of sufficiently high degree, in a given graph. This node does not necessarily have the greatest degree in the graph, and often we cannot be sure whether it does as the greatest degree is unknown beforehand. This problem can be formally stated as

>>>>

\usepackage[margin=1in, paperwidth=8.3in, paperheight=11.7in]{geometry}

\usepackage[ruled, vlined, linesnumbered]{algorithm2e}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{changepage}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{fancyhdr} % Header
\usepackage{float} % [H] for figure
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{tikz}

@@

\renewcommand{\headrulewidth}{0pt}
\newcommand{\ie}{\textit{i.e.} }
\newcommand{\nats}{\mathbb{N} }
\newcommand{\horizontalline}{\newline\vspace{.3cm}\hfill\makebox[.5\linewidth]{\rule{.5\textwidth}{0.4pt}}\hfill\vspace{.05cm}}

\newcommand{\smallsim}{\smallsym{\mathrel}{\sim}}

\makeatletter
\newcommand{\smallsym}[2]{#1{\mathpalette\make@small@sym{#2}}}
\newcommand{\make@small@sym}[2]{%
  \vcenter{\hbox{$\m@th\downgrade@style#1#2$}}%
}
\newcommand{\downgrade@style}[1]{%
  \ifx#1\displaystyle\scriptstyle\else
    \ifx#1\textstyle\scriptstyle\else
      \scriptscriptstyle
  \fi\fi
}
\makeatother

\renewcommand\thetable{\arabic{table}} % figure numbering strategy
\renewcommand{\thefigure}{\arabic{chapter}.\arabic{section}.\arabic{figure}}

\pagenumbering{roman}

% Header
\pagestyle{fancy}
\fancyhead[L]{Dom Hutchinson (170 1111)}
\fancyhead[C]{}
\fancyhead[R]{\today}

% Algorithm environment
\SetArgSty{textnormal}
\DontPrintSemicolon

% Title
\title{Implementing \& Evaluating Space Efficient Algorithms for Detecting Large Neighbourhoods in Graph Streams}
\author{Dom Hutchinson}
\date{\today}
\maketitle
\newpage

\chapter*{Declaration}

\tableofcontents
\newpage

\chapter*{Abstract}

\par The \texttt{Neighbourhood Detection Problem} is a problem in graph theory which tasks one with finding a vertex in the graph which has a certain number of neighbours and to then return a subset of these neighbours. This is a trivial problem to solve in theory, but in practice is hard to solve within a time frame which makes the implementation practical. A major reason for this is that many approaches to solving the problem require a lot of space and so quickly overflow a computer's RAM, slowing the execution significantly down. This common space inefficiency is becoming increasingly problematic since the move towards Big Data.

\par In this paper I discuss and implement the two algorithms presented in \cite{orig} which offer space efficient solutions for \texttt{neighbourhood detection} for two types of graph stream. During the implementation I discuss and evaluate several methods for improving performance of these algorithms, focusing on space-efficiency in order to maximise the potential size of graphs the implementations are effective on.

\par The final implementation of the proposed algorithm for insertion-only streams proves very promising with it able to find large neighbourhoods in a couple of minutes for a graph with 30 million edges. The algorithm for insertion-deletion streams proves much more difficult to evaluate due to restrictions incurred from the implementation of $L_0$ samplers.

\chapter*{Acknowledgments}

\renewcommand\thechapter{\Roman{chapter}}
\renewcommand\thesection{\thechapter.\roman{section}}
\setcounter{chapter}{1}
\chapter*{Introduction}
\addcontentsline{toc}{chapter}{Introduction}

@@

%\begin{itemize}
%	\item High Degree Detection.
%	\item Large Neighbourhood Detection.
%\end{itemize}
Graphs are used to represent relationships between objects and are popular in data science. A graph stream is a sequence of instructions which describe how to construct a graph. The sequential nature of a graph stream allows it to represent the evolution of a graph over time, which adjacency matrices cannot. This makes graph streams ideal for representing changing networks such as a social or computer network where objects can connect and then disconnect.

\par A basic problem in graph theory is \texttt{High Degree Detection} where you want to find the vertices with the greatest degree. This problem can be trivially and efficiently solved by counting the number of edges incident to each vertex and then returning any vertices of sufficient degree. Solving \texttt{High Degree Detection} allows for the identification of the most influential vertices in a graph, however you cannot make any inferences about the nature of this influence. In a social network it is possible for someone to have lots of followers (\ie be of high degree), but for all these followers to be bot accounts meaning the account actually has zero influence.

\vspace{.3cm}\begin{adjustwidth}{.3cm}{}\fbox{\parbox{\textwidth}{
\textbf{Problem 1} \texttt{Neighbourhood Detection}.\\
Let $G=(A\cup B,E)$ be a bi-partite graph with vertex sets $A,B$, where $|A|=n$ and $|B|=\text{poly }n$, and edge-set $E$.\\
In $\texttt{Neighbourhood Detection}(G,d,c)$ we are tasked with outputting a vertex from $A$ with at least $d/c$ of its neighbours in $B$. We can assume that $G$ contains at least one node of degree $d$.\\
Here $d\in\nats$ is a threshold parameter \& $c\in(0,1]$ is an approximation parameter.
}}\end{adjustwidth}\vspace{.3cm}

\par\texttt{Neighbourhood Detection} is the natural set up from \texttt{High Degree Detection}, it tasks you with finding vertices of high degree \underline{and} a subset of their neighbours. Solving \texttt{Neighbourhood Detection} allows for inferences to be made about the influence a vertex has. Extending the social network example, \texttt{Neighbourhood Detection} allows for analysis of the followers of a popular account which can be used for targeted marketing campaigns.

\par\texttt{Neighbourhood Detection} can be solved trivially by storing the neighbourhood of every vertex and then returning any subset of size $\frac{d}c$. This trivial solution is very space inefficient and, with the movement towards Big Data, would quickly overflow a computer's RAM. This results in the use of virtual data which significantly slows down data access time and thus the whole algorithm. This motivates the need for space efficient solutions to \texttt{Neighbourhood Detection}.

\section{Motivating Applications} %TODO I don't really like this (partly because it is in list form), moreover it isn't really connected in
% NOTE maybe it is best that it is fairly separate from the rest so that it can be skipped over.
%\begin{itemize}
%	\item Real world occurrences \& uses.
%\end{itemize}

There are several real-world applications of \texttt{Neighbourhood Detection} which motivate its investigation. These are situations where simply knowing highly connected elements of a network is not sufficient.
\begin{itemize}
	\item[-] Given a list of connections in a social network, identify \textit{social influencers} and determine the demographics of their audience in order to decide who should promote a particular product.
	\item[-] Given a list of receipts, identify which items are commonly sold together and use this information to determine how to stack shelves in order to increase sales.
	\item[-] Given a log of traffic within a network identify which resources are being accessed most often, and by whom. This information can be used to determine what upgrades should be made to the network and to identify potential attacks on the network.
\end{itemize}

\section{Objectives}
The objectives of this project are to:
\begin{itemize}
	\item Implement and understand the algorithms proposed in \cite{orig}.
  \item Discuss and test alterations to the logic and implementation of the proposed algorithms, and the subroutines they rely upon, in order to improve the performance of the proposed algorithm.
	\item Use large graphs from real-world scenarios to evaluate the practicalities of these implementations on real-world data sets.
\end{itemize}

\section{Structure}
Having provide some background to the problem, the structure of this project is as follows. In \texttt{Chapter 1} I discuss and define general ideas, technologies and techniques which are used throughout this project. In \texttt{Chapter 2} I discuss, implement and evaluate the algorithm proposed in \cite{orig} for insertion-only graph streams. In \texttt{Chapter 3} I discuss, implement and evaluate the algorithm proposed in \cite{orig} for insertion-deletion graph streams. And, in \texttt{Chapter 4} review what was achieved in this project and provide some ideas for possible future work to build upon those achievements.

\renewcommand\thechapter{\arabic{chapter}}
\renewcommand\thesection{\thechapter.\arabic{section}}
\setcounter{chapter}{0}
\chapter{Preliminaries}

\setcounter{page}{1} % restart page counter
\pagenumbering{arabic}

\section{Definitions}

%\begin{itemize}
%	\item Graph
% \item Unweighted, undirected
%	\item Bipartite Graph (Why this wasn't really applied in practice, A=B)
%	\item Graph Stream (Insertion-Only, Insertion-Deletion)
%	\item Neighbourhood, degree & sparsity of a vertex
%	\item Streaming Algorithm
%	\item $1-\varepsilon$ Approximation Algorithm
%	\item TODO Space efficiency??
%\end{itemize}

A \textit{Graph} is a data structure used to represent pairwise relationships between objects. A Graph $G=(V,E)$ is defined to have a vertex set $V$, which holds the objects, and an edge-set $E$, which holds the relationship between the objects. Graphs are traditionally visualised with circles for each vertex and lines between vertices that share an edge. There are variations on graphs which allow for edges to have direction and weight. In this project we are using undirected, unweighted graphs meaning edges can be represented by an unordered pair of vertices $(u,v)$ for $u,v\in V$.

\par A graph $G=(V,E)$ is said to be a \textit{Bipartite Graph} if its vertex set $V$ can be partitioned into two disjoint subsets $A,B$ and the every edges connects a vertex in $A$ to a vertex in $B$. Formally, $\exists\ A,B\subseteq V$ st $A\cup B=V,\ A\cap B=\emptyset$ and $\forall\ (a,b)\in E$ we have that $a\in A,b\in B$. \textit{Bipartite Graphs} are denoted as $G=(A\cup B,E)$.

\begin{figure}[H]
	\centering
	\subfloat[Graph]{{\begin{tikzpicture}
		\node[shape=circle] (v) at (0,0) {$v$};
		\node[shape=circle] (w) at (0,-1) {$w$};
		\node[shape=circle] (x) at (1,0) {$x$};
		\node[shape=circle] (y) at (1,-1) {$y$};
		\node[shape=circle] (z) at (2,-1) {$z$};

		\path [-] (v) edge (w);
		\path [-] (x) edge (y);
		\path [-] (x) edge (z);
	\end{tikzpicture}}}
	\qquad
	\subfloat[Bipartite Representation]{{\begin{tikzpicture}
		\node[shape=circle] (A) at (-1,0) {$A$};
		\node[shape=circle] (v) at (0,0) {$v$};
		\node[shape=circle] (w) at (1,0) {$w$};
		\node[shape=circle] (x) at (2,0) {$x$};
		\node[shape=circle] (y) at (3,0) {$y$};
		\node[shape=circle] (z) at (4,0) {$z$};
		\node[shape=circle] (B) at (-1,-1) {$B$};
		\node[shape=circle] (v') at (0,-1) {$v'$};
		\node[shape=circle] (w') at (1,-1) {$w'$};
		\node[shape=circle] (x') at (2,-1) {$x'$};
		\node[shape=circle] (y') at (3,-1) {$y'$};
		\node[shape=circle] (z') at (4,-1) {$z'$};

		\path [-] (v) edge (w');
		\path [-] (x) edge (y');
		\path [-] (x) edge (z');
		\path [-] (v') edge (w);
		\path [-] (x') edge (y);
		\path [-] (x') edge (z);
	\end{tikzpicture}}}
	\caption{Example of how all graphs have a bipartite representation}
	\label{Figure 0}
\end{figure}

\par All graphs $G=(V,E)$ have a bipartite representation $G'=(A\cup B,E')$. Considering the example in \texttt{Figure 1}, defining $A=V$, $B$ to a be a copy of $V$, and the edge set \newline${E'=\{(u,v'):(u,v)\in E,u\in A,v'\in B\}}$ generates a bipartite presentation of $G$. The bipartite representation has twice the number of edges since each edge $(u,v)$ is now replaced by two edges $(u,v')$ and $(v,u')$. %TODO in practice we don't need these duplicates

\par The \textit{Neighbourhood} of a vertex $v$ is the set of vertices which share an edge with, $N_v=\{u:(u,v)\in E\}$. The \textit{Degree} of a vertex $v$ is the size of its neighbourhood, $\delta_v=|N_v|$. A vertex $v$ is \textit{$s$-sparse} if its degree is less than, or equal to, $s$.
\par The \textit{Edge Vector} of a vertex $v$ of length $|V|$ where each index gives the weight of the edge from $v$ to the vertex associated with that index. Typically a weight of 0 is stored if no edge exists between $v$ and a specific vertex. The graphs used in this project are unweighted so the values in the edge vector are booleans describing whether, or not, an edge exists. These edge vectors can be stored as bitstrings.
\par \textit{Graph Streams} are an unbounded sequence of instructions $\{e_1,e_2,\dots\}$ which describe how to construct the graph. These instructions $e_i$ describe modifications to the edges of a graph. The vertex set of the described graph is inferred from the instructions. In this project we are only concerned with two types of graph stream. %TODO something about the unbounded part??
\begin{itemize}
	\item \textit{Insertion-Only Streams} where each instruction inserts a new edge to the graph. It is assumed that no edges are inserted multiple times. For the unweighted, undirected graphs used in this project insertion-only streams are a list of the edges in the graph and can be read in any order. Here the instructions have the form $e_i=(u,v)$ where $u,v\in V$ are the endpoints of the edge being inserted.
	\item \textit{Insertion-Deletion Streams} where each instruction either adds a new edge to the graph or removes an existing edge from the graph. It is assumed that no edges are inserted multiple times, nor is an edge removed if it is not currently in the graph. For the unweighted, undirected graphs used in this project instructions of insertion-deletion streams take the form $e_i=(\Delta,u,v)$ where $u,v\in V$ are the endpoints of the edge and $\Delta$ is a boolean defining whether $e_i$ is an insertion or deletion instruction. The order of instructions is clearly important for insertion-deletion streams.
\end{itemize}
Other versions of graph streams exist which allow for changes to the weights of edges and directed edges, but these are out of scope for this project.

\par \textit{Streaming Algorithms} are algorithms which are designed to take a stream of sequential instructions as their input. Streaming algorithms are space-efficient since only one instruction is read at a time, requiring constant space. Streaming algorithms which only require a single pass of the instruction stream are called \textit{Online Algorithms}. Online algorithms have the advantage of being able to take new instructions without having to recompute anything.

\par An \textit{$\alpha$-Approximation Algorithm} is an algorithm which returns a result within an $\alpha$ factor to the optimal solution to the problem. Algorithms which solve the \texttt{Neighbourhood Detection Problem} are $\frac1c$-approximation algorithms since they only return $d/c$ of the neighbours of a vertex which we know to have at least $d$ neighbours.

\par A family of hash functions $\mathcal{H}=\{h:X\to Y\}$ is described as being \textit{Pairwise Independent} if $\forall\ u,v\in X$ with $u\neq v$ and $\forall\ a,b\in Y$ if $h$ is chosen uniformly at random from $\mathcal{H}$ then $\mathbb{P}(h(u)=a\wedge h(v)=b)=\frac1{|Y|^2}$. This means that the values $u,v$ are hashed to are assigned uniformly at random and pairwise independently.

\par A family of hash functions $\mathcal{H}=\{h:X\to Y\}$ is described as begin \textit{$k$-Wise Independent} if for any $k$ distinct elements of $X$ $\{v_1,\dots,v_k\}$ and any, not necessarily distinct, elements of $Y$ $\{a_1,\dots,a_k\}$ if $h$ is chosen uniformly at random from $\mathcal{H}$ then $$\mathbb{P}(h(v_1)=a_1\wedge\dots\wedge h(v_k)=a_k)=\frac1{|Y|^2}$$
\section{Technologies}
%\begin{itemize}
%	\item C++
%	\begin{itemize}
%		\item Low level
%		\item Memory manipulation
%		\item Faster
%		\item Perhaps something about why not java or python etc.
%	\end{itemize}
%	\item Standard library online
%\end{itemize}
For the implementations in this project I chose to use C++, specifically \texttt{C++11}. I initially considered using either python or C++, as they both work in the object-orientated paradigm and I am proficient in both. I chose C++ as it allows for more control with memory management and typically has faster run-times. For this project the memory management is more important as I am looking to evaluate the space-efficiency of algorithms. The faster run-times is important when evaluating the practicalities of the algorithms.
\par I limited myself to using the \textit{C++ Standard Library} \texttt{std} as the underlying implementations of this library are well document. This was particularly important for the abstract data types I used as I could check what the underlying implementations and adjust my evaluation accordingly. When implementing randomness I used the \texttt{<random>} module of \texttt{std} and seeded the generators with the current time so that each run would have a different generator.

\section{Evaluation Approach}

% TODO expectations
%\begin{itemize}
%	\item What features are tested (time \& space)? How?
%	\item Against naïve algorithm.
%	\item Insertion-Only v Insertion-Deletion algorithm for insertion-only streams.
%\end{itemize}

%TODO talk about verifying my results
%TODO typical test = vary c between [3,20]. set d=max degree

During evaluation three performance metrics were measured: success-rate, time-taken, and, space-used. These are all interconnected and come with their own trade-offs. Typically, the changes to space-used \& time-taken when different strategies are implemented occur in the same direction, but at different rates. During most tests the success-rate was near perfect, but the other metrics were significantly higher as it took longer for a solution to be found.

\par In real world scenarios we are most interested analysing the neighbourhoods of the most influential vertices. For this reason $d$ is set to the maximum degree in the graph during all tests. Further, the greater the proportion of a neighbourhood returned the better the inferences made are so I focussed testing on low values of $c$, namely $c\in[1,20]$.

\par For this project I wanted some graphs generated from the real-world. I found a collection of large graph streams, from different applications, in the \texttt{Stanford Network Analysis Project} (\texttt{SNAP}) datasets collection \cite{SNAP}. I chose to use the graphs generated from social networks as I found that application most motivating, but this choice is essentially arbitrary. The graph streams from \cite{SNAP} were all insertion-only so I wrote a utility program which created insertion-deletion streams from them by using a Bernoulli random variable to decide whether to delete an edge immediately after inserting it. \textbf{Table 1} provides details about the graphs used during evaluation.

\begin{center}
	\begin{table}[h]
		\tiny
		\begin{tabular}{|l|l|l|l|l|l|l|l|}
			\hline
			\textbf{Name}&\textbf{Type}&\textbf{\# Vertices}&\textbf{\# Edges}&\textbf{\# Instructions}&\textbf{Max Degree}&\textbf{File Size}\\
			\hline
			$\mathtt{facebook\_small}$&Insertion-Only&52&146&-&18&3 KB\\
			$\mathtt{facebook\_small\_deletion}$&Insertion-Deletion&52&131&161&16&5 KB\\
			$\mathtt{facebook}$&Insertion-Only&747&30,025&-&293&587 KB\\
			$\mathtt{facebook\_deletion}$&Insertion-Deletion&747&26,718&33,332&267&846 KB\\
			$\mathtt{gplus}$&Insertion-Only&12,417&1,179,613&-&5,948&12 MB\\
			$\mathtt{gplus\_deletion}$&Insertion-Deletion&12,417&1,049,309&1,309,917&4,998&16 MB\\
			$\mathtt{gplus\_large}$&Insertion-Only&102,100&30,238,035&-&104,947&1.3 GB\\
			\hline
		\end{tabular}
    \caption{Details of graphs used during evaluation}
	\end{table}
\end{center}

\chapter{Insertion-Only Streams}

\section{Degree-Based Reservoir Sampling}

\texttt{Degree-Based Reservoir Sampling} is a technique for uniformly sampling a vertex from the set of vertices whose degrees are greater than a specified minimum bound $d_1$ along with $d_2$ of its neighbours. This is achieved by maintaining a sample of this set of vertices, known as the \textit{reservoir}, and storing the first $d_2$ of the edges incident to a vertex after it is sampled. The size of the reservoir is a parameter of the algorithm. \textbf{Algorithm 1} outlines pseudocode for performing \texttt{Degree-Based Reservoir Sampling} on a bipartite graph using an insertion-only graph stream.

\begin{algorithm}
	\caption{\texttt{Degree-Based Reservoir Sampling}$(d_1,d_2,s)$}
	\SetKwInOut{Require}{require}
	\Require{Degree bound $d_1\in\nats$, Neighbourhood bound $d_2\in\nats$, Reservoir size $s\in\nats$, Insertion-only stream $\{(a_0,b_0),\dots,(a_n,b_n)\}$}
	$D\leftarrow\{\}$ \tcp*{Degree counter}
	$R\leftarrow\{\}$ \tcp*{Reservoir}
	$E\leftarrow\{\}$ \tcp*{Collected edges}
	$x\leftarrow0$\tcp*{\# nodes of degree$\geq d_1$}
	\For{$i\in[0,n]$} {
		$D[a_i]\leftarrow D[a_i]+1$\tcp*{Increment degree counter}
		\If{$D[a_i]\equiv d_1$}{
			\tcp{Consider inserting $a_i$ to reservoir}
			$x\leftarrow x+1$\\
			\If{$|R|<s$} {
				\tcp{Reservoir is not full}
				$R\leftarrow R\cup\{a_i\}$
			} \Else {
				\tcp{Reservoir is full}
				\If{\texttt{Bernoulli}$\left(\frac{s}x\right)$} {
					Let $a'$ be a uniform random element of $R$\tcp*{Element to replace}
					Delete edges in $E$ incident to $a'$\\
					$R\leftarrow(R\backslash\{a'\})\cup\{a\}$\tcp*{Swap $a_i$ and $a'$}
				}
			}
		}
		\If{$a_i\in R$ \textbf{and} $D[a_i]< d_1+d_2$}{
			$E\leftarrow E\cup(a_i,b_i)$\tcp*{Store edge}
		}
	}
	\If{$\exists\ a\in R$ with $D[a]\geq d_1+d_2-1$} {
		\Return{Uniform random vertex \& neighbourhood from those of size $d_2$}
	} \lElse { \Return{\texttt{FAIL}}}
\end{algorithm}

\par Three data stores are used in \texttt{Degree-Based Reservoir Sampling}: a map of the degree of every node in the graph $D$; a set for the reservoir $R$; and a set for edges incident to the vertices in the reservoir $E$.

\par The reservoir has an invariant that at any point in time it contains a uniform sample of the vertices whose degrees are known to be greater than $d_1$. This invariant is proved as \textbf{Lemma 2.1}, first I shall describe how the reservoir is maintained by controlling how vertices are inserted. The first time the degree counter for a vertex $v$ surpasses $d_1$, $v$ is considered for insertion into the \textit{reservoir}. There are two cases
\begin{itemize}
	\item If the reservoir is \underline{not} full (\ie $|R|<s$) then $v$ is inserted into the reservoir.
	\item Otherwise, if the reservoir is already full, a Bernoulli random variable with probability $p=\frac{x}s$ is used, where $x$ is the number of vertices known to have degree at least $d_1$. If this random variable succeeds then pick, uniformly at random, a vertex $u$ currently in the reservoir and replace it with $v$. All the edges in $E$ which are incident to $u$, and no other vertex in the reservoir, are removed from $E$.
\end{itemize}

Whenever an edge $(u,v)$ is encountered and $u$ is in the reservoir and the number of edges stored in $E$ incident to $u$ is less than $d_2$ then $(u,v)$ is inserted into $(u,v)$. %TODO rewrite this
\begin{adjustwidth}{.3cm}{}\fbox{\parbox{\textwidth}{

\textbf{Lemma 2.1} At any time $t$, $R$ contains a uniform sample of the vertices whose degrees are known to be at least $d_1$.\\

\textit{Proof (by induction)}\\
Let $X_t$ be the set of vertices whose degrees are know to be greater than $d_1$ after $t$ instructions have been consumed and $x_t:=|X_t|$.\\
\underline{Base Case} - $x_t\leq s$.\\
Trivially true since $R$ is not yet full so $R=X_t$.\\
\underline{Inductive Case} - $x_t>s$.\\
Assume the property holds for $x_t=z>s$, then the probability of any given element of $X_t$ is in $R$ is $\frac1z$.
Without loss of generality, let $a$ be the next element considered for sampling and $b$ be any element of $R$. The probability $a$ is sampled is $\frac{s}{z+1}$ and, given $a$ is to be sampled, the probability $b$ is chosen for removal is $\frac1s$. Thus, the probability that $b$ is replaced by $a$ is $\frac{s}{z+1}\cdot\frac{1}{s}=\frac1{z+1}$. Similarly, the probability $b$ is \underline{not} replaced by $a$ is $1-\frac1{z+1}=\frac{z}{z+1}$. By the inductive hypothesis, the probability that $b$ was in $R$ was $\frac1z$. Thus, the probability that $b$ is still in $R$ is $\frac1z\times\frac{z}{z+1}=\frac1{z+1}$.\\

Hence by mathematical induction, $R$ contains a uniform sample of the vertices of sufficient degree at all points in time.$\hfill\square$
}}\end{adjustwidth}\vspace{.3cm}
%TODO adding termination condition? Maybe add during implementation

\par If $v$ is not added to the reservoir at this time, it never will be in the future. Note that in the case that the reservoir is full and $v$ is inserted into it, $v$ is replacing an element that was in the reservoir meaning any progress made towards finding a neighbourhood for that vertex is annulled. The amount of progress is not taken into account \& thus if there is a high turn-over of elements in a reservoir then less progress is made towards finding a neighbourhood for each of them. This is discussed further during evaluation.

\par \texttt{Degree-Based Reservoir Sampling} only works on insertion-only streams as it does not account for when the degree of a vertex falls below the threshold, after being sampled. This means the reservoir is not necessarily a uniform random sample of the vertices with degree $\geq d_1$. %TODO is this sufficient

\par For this project we are interested in how \texttt{Degree-Based Reservoir Sampling} can be implemented for bipartite graphs, specifically we want to sample from the $A$-vertices of the graph. For a bi-partite graph with $n$ $A$-vertices \texttt{Degree-Based Reservoir Sampling} requires $\mathcal{O}(n\log n+sd_2\log n)$ space, assuming $O(\log n)$ space is required to store an edge or a vertex, since it stores a degree counter for every $A$-vertex and at most $d_2$ edges for each of the $s$ vertices in the reservoir.

\section{Proposed Algorithm}

%\begin{itemize}
%	\item Assumptions: no duplicate edges (can mitigate cases when this would be required by being smart)
%	\item Insertion stream neighbourhoods cannot be diminished over time.
%	\item Possibility of failure (missed edges) \& mitigation
%	\begin{itemize}
%		\item Theorem 3.2 (probs not proof)
%	\end{itemize}
%	\item Theoretical space requirements
%\end{itemize}

The \texttt{Neighbourhood-Detection Problem} can be solved using \texttt{degree-based reservoir sampling} sampling by setting $d_2=\frac{d}c$. The question remains as to what to set $d_1$ and $s$ to in order to achieve a high probability of success. %in little time and space

\par \textbf{Algorithm 2} is the algorithm proposed in \cite{orig} for solving the \texttt{Neighbourhood-Detection Problem} for insertion-only streams.

\begin{algorithm}
	\caption{One-pass $c$-Approximation Insertion-Only Streaming Algorithm for $\mathtt{Neighbourhood\ Detection}$}
	\SetKwInOut{Require}{require}
	\Require{Space $s$, degree bound $d$.}
	$s\leftarrow\lceil\log(n)\cdot n^{\frac1c}\rceil$\\
	\For{$i\in[0,c-1]$\text{\textbf{ in parallel}}} {
		$(a_i,S_i)\leftarrow\mathtt{Deg}\mbox{-}\mathtt{Res}\mbox{-}\mathtt{Sampling}\left(\max\left\{1,i\cdot\frac{d}{c}\right\},\frac{d}c,s\right)$
	}
	\Return{\small Uniform random neighbourhood $(a_i,S_i)$ \text{from successful runs}}
\end{algorithm}

\par In \cite{orig} it is proven that setting $s=\lceil\ln(n)\cdot n^{\frac1c}\rceil$ and using $c$ \texttt{degree-based reservoir samplers} each with a different lower bound, incremented from $\frac{d}c$ to $d$ stepping by $\frac{d}c$ each time, results in a high probability of success. The samplers are run in parallel so that the algorithm only requires a single pass of the stream. Implementing parallel running of the samplers just requires passing the same instruction to each sample before fetching the next instruction.

\par Each of the $c$ samplers requires $\mathcal{O}(n\log n+sd_2\log n)$ space. Since the degree map can be shared between samplers the total space is $\mathcal{O}(n\log n+c\cdot sd_2\log n)=\mathcal{O}(n\log n+c\lceil\log(n)n^{\frac1c}\rceil\frac{d}c\log n)=\mathcal{O}(n\log n+n^{\frac1c}d\log^2n)$. The proposed algorithm does require you to know the number of vertices in the graph stream before running the algorithm, or at least the $\log$ of it. If this is unknown then it is quick to run through the whole stream \& build a set of vertices in the graph. This requires $\mathcal{O}(n\log n)$ space, assuming $O(\log n)$ space is required to store a vertex. However, in many real-world scenarios this number (or a good approximation) will be known due to it being important for other tasks.
